---
title: "Reproducible Research: Peer Assessment 1"
output: github_document
---
 
## Loading and preprocessing the data
#### Unzip archive and read data
```{r echo=TRUE}
# graphics library load
library(ggplot2)

# Unzip archive
unzip("activity.zip")

# Read base data into a data frame.
baseData <- read.csv("activity.csv")
```

#### Summarize Data
```{r  echo=TRUE}
# Data Head
head(baseData)

# Data Dimensions
dim(baseData)

# Data Summary
summary(baseData)
```

#### NA and Date String to Date Class modification
```{r echo=TRUE}
# change date type to date class 
baseData$date <- as.Date(as.character(baseData$date))
# create NA logical vector
baseDataNA <- is.na(baseData$steps)
# create clean base for later mean calculations
cleanBase <- baseData[!baseDataNA,]
```

## What is mean total number of steps taken per day?
#### 1. Calculate the total number of steps taken per day
```{r echo=TRUE}
# aggregate clean non NA steps per day (SUM)
SummedDataByDay <- aggregate(baseData$steps, by=list(baseData$date), sum)
# adjust column names
names(SummedDataByDay)[1] ="date"
names(SummedDataByDay)[2] ="totalsteps"
# top 15 of Summed Steps by day
head(SummedDataByDay,15)
```
#### 2. Make a histogram of the total number of steps taken each day
```{r echo=TRUE}
# Plot using ggplot
ggplot(SummedDataByDay, aes(x = totalsteps)) +
  geom_histogram(fill = "steelblue", binwidth=1000) +
  labs(title = "Total Daily Steps", x = "Steps", y = "Frequency")
```

#### 3. Calculate and report the mean and median of the total number of steps taken per day
```{r echo=TRUE}
# Mean of steps taken per day
mean(SummedDataByDay$totalsteps,na.rm=TRUE)
#median of steps taken per day
median(SummedDataByDay$totalsteps,na.rm=TRUE)

```

## What is the average daily activity pattern?
#### 1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r echo=TRUE}
nonNASubset <- baseData[!baseDataNA,]
MeanDataByInterval <- aggregate(nonNASubset$steps, by=list(nonNASubset$interval), mean)
# set the column names
names(MeanDataByInterval)[1] ="interval"
names(MeanDataByInterval)[2] ="steps"

ggplot(MeanDataByInterval, aes(x = interval, y=steps)) +
  labs(title = "Sum of Steps by Interval", x = "interval", y = "steps")+
  geom_line(color="red") 
```

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r echo=TRUE}
maxInterval <- MeanDataByInterval[which.max(MeanDataByInterval$steps),]
maxInterval
```
 
## Imputing missing values
#### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r echo=TRUE}
# generate listing of NA's
missingVals <- sum(baseDataNA)
missingVals
```
#### 2. Devise a strategy for filling in all of the missing values in the dataset. 
#### Strategy: Use mean interval steps from Mean Steps for that interval.
#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r echo=TRUE}
# new dataset
baseData2 <- baseData
# NA's in dataset
# dataset minus NA's for Mean calculation
NABase2 <- baseData2[is.na(baseData2$steps),]
cleanBase2 <- baseData2[!is.na(baseData2$steps),]

# generate Mean Data2 by interval
MeanData2ByInterval <- aggregate(cleanBase2$steps, by=list(cleanBase2$interval), sum)
names(MeanData2ByInterval)[1] ="interval"
names(MeanData2ByInterval)[2] ="steps"

# IMPUT METHOD- ATTEMPT 2
baseData2 <- baseData
missingData <- is.na(baseData2$steps)
meanVals <- tapply(cleanBase$steps, cleanBase$interval, mean, na.rm=TRUE, simplify=TRUE)
baseData2$steps[missingData] <- meanVals[as.character(baseData2$interval[missingData])]

#IMPUTE METHOD ATTEMPT 1.
# after trying the following code and messing with it for a few hours, and couldn't quite 
# get it right, i abandoned it for another path.  I expect the assignment was slightly off.
# intervals to match
#myCount <- 0
# loop through baseData2
#for (i in 1:nrow(baseData2)){
#  # if steps in current row is null act.
#  if (is.na(baseData2$steps[i])){
#    # match interval
#    for (j in 1:length(MeanData2ByInterval$interval)){
#      if (baseData2$interval[i]== MeanData2ByInterval$interval[j]){
#        baseData2$steps[i] <- as.integer(MeanData2ByInterval$steps[j])
#        myCount <- myCount + 1
#      }
#    }
#  }
#}
# values modified.
#myCount 

# original missing data count
sum(missingData)
# count of NA values
sum(is.na(baseData2$steps))
```
#### 4a) Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 
```{r echo=TRUE}
FullSummedDataByDay <- aggregate(baseData2$steps, by=list(baseData2$date), sum)

names(FullSummedDataByDay)[1] ="date"
names(FullSummedDataByDay)[2] ="totalsteps"
head(FullSummedDataByDay,15)
# Plot using ggplot
ggplot(FullSummedDataByDay, aes(x = totalsteps)) +
  geom_histogram(fill = "steelblue", binwidth=1000) +
  labs(title = "Total Daily Steps", x = "Steps", y = "Frequency")

# Mean on New Data
mean(FullSummedDataByDay$totalsteps)
# Median on New Data
median(FullSummedDataByDay$totalsteps)
```

#### 4b) Do these values differ from the estimates from the first part of the assignment? 
#### Yes, he mean is the same but the median has risen 1.19 steps.
#### Original Mean and Median
```{r echo=FALSE}
mean(SummedDataByDay$totalsteps,na.rm=TRUE)
median(SummedDataByDay$totalsteps,na.rm=TRUE)
```
#### New Mean and Median
```{r echo=FALSE}
mean(FullSummedDataByDay$totalsteps)
median(FullSummedDataByDay$totalsteps)
```
#### 4c) What is the impact of imputing missing data on the estimates of the total daily number of steps?
#### The effect of using mean data per interval as a data impute method for missing values seems to push overall data towards the mean.

## Are there differences in activity patterns between weekdays and weekends?
```{r echo=TRUE}
baseData2$weekday <- weekdays(baseData2$date)
baseData2$weekend <- ifelse (baseData2$weekday == "Saturday" | baseData2$weekday == "Sunday", "Weekend", "Weekday")
#baseData2$weekend <- as.factor(baseData2$weekend)
head(baseData2,5)
MeanDataWeekendWeekday <- aggregate(baseData2$steps, by=list(baseData2$weekend, baseData2$interval), mean)
names(MeanDataWeekendWeekday)[1] ="weekend"
names(MeanDataWeekendWeekday)[2] ="interval"
names(MeanDataWeekendWeekday)[3] ="steps"

ggplot(MeanDataWeekendWeekday, aes(x = interval, y=steps, color=weekend)) +
  geom_line() +
  facet_grid(weekend ~ .) +
  labs(title = "Mean of Steps by Interval", x = "interval", y = "steps")
```

#### There seems to be variation in the beginning of the day during weekdays, likely due to workplace activities.  There seems to be an overall slightly larger incidence of steps during the weekends.
